---
title: "Movielens edx-PH125.9x"
author: "Jin Xin"
date: "Sep, 2022"
output: pdf_document
bibliography: references.bib
---

an introduction/overview/executive summary section that describes the dataset and summarizes the goal of the project and key steps that were performed

a methods/analysis section that explains the process and techniques used, including data cleaning, data exploration and visualization, insights gained, and your modeling approach

a results section that presents the modeling results and discusses the model performance

a conclusion section that gives a brief summary of the report, its limitations and future work

40 points: The report includes all required sections, is easy to follow with good supporting detail throughout, and is insightful and innovative.
## Executive summary
We are going to create a movie recommendation system in this project. This system will predict the possible rating a movie likely to receive from the specified user. We use the RMSE(root mean square error) to evaluate our algorithm performance, and the final achieved RMSE is \textcolor{blue}{?0.80}. The algorithm is based on the Netflix winning latent factor model [@Koren2009], but a rather vanilla version. 

## Introduction
This project is inspired by the $1 million prize competition hosted by Netflix in 2006. The rule of winning is to achieve a 10% more improvement on RMSE, comparing to Netflix own in house algorithm. But in this project, we just use a fraction of Netflix competition dataset (our 10m version against the original size of 100m [@Töscher09]). Our goal in here is simply to achieve a \textcolor{blue}{RMSE < 0.86490} between our prediction and validation set ratings.

In the following sessions, we will look into the dataset first to have a macro view, then we will explain details of the methods to achieve the final goal. 

## Data exploration

First of all, we use the following code to download and prepare the data we need. The code in below is provided by edx-HarvardX PH125.9x. To successfully obtain the `edx` and `validation` dataset, we need to follow the instructions inside the code chunk below accordingly, with respect to our own computer setup.  

```{r eval=FALSE}
##########################################################
# Create edx set, validation set (final hold-out test set)
##########################################################

# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", 
                            repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", 
                            repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", 
                            repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("https://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", 
                             readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), 
                          "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

# if using R 3.6 or earlier:
movies <- as.data.frame(movies) %>% 
  mutate(movieId = as.numeric(levels(movieId))[movieId],
                                            title = as.character(title),
                                            genres = as.character(genres))
# if using R 4.0 or later:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                            title = as.character(title),
                                            genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
# if using R 3.5 or earlier, use `set.seed(1)`
set.seed(1, sample.kind="Rounding") 
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, 
                                  list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
      semi_join(edx, by = "movieId") %>%
      semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

We will use the `edx` data to develop and train our model, then apply the trained model on the `validation` data to predict the rating with respect to the user/movie pairs, and evaluate the prediction against the real ratings by the **RMSE** method. So let's look at the macro structure of the dataset first. 

```{r include = FALSE}
# these libraries are used
library(tidyverse)
library(data.table)
library(caret)
library(Matrix)
library(doParallel)
library(glmnet)
library(RcppArmadillo)
load("~/Work/Rproj/PH125.9x/harvardx-ds-capstone/misc/edx.Rdata")
load("~/Work/Rproj/PH125.9x/harvardx-ds-capstone/misc/val.Rdata")
```

Here is the macro view of `edx` dateset:
```{r echo = FALSE}
options(width = 70)

edx %>% tibble()
```

Here is the macro view of `validation` dataset:
```{r echo = FALSE}
options(width = 70)

validation %>% tibble()
```

Comparing the `edx` and `validation` dataset, we can tell the `edx` is just a longer version of `validation` with different user/movie pairs. When we look into the details of 'edx' dataset, we have:
```{r echo = FALSE}
options(width = 70)

str(edx)
```

In the 6 data frame headers, we can classify them into three groups: user info, movie info, and rating (the joint info). 

For user info, we have 'userId', 'timestamp'. This 'timestamp' is showing the information about when the movie is rated by the user, and in the (Unix Epoch Time)[https://files.grouplens.org/datasets/movielens/ml-10m-README.html] format. 

For movie info, we have 'movieId', 'title', and 'genres'. One thing need to pay attention is that, one movie can be categarised as multiple 'genres'. 

let's look at one example to have a more straight understanding. 
```{r echo = FALSE}
options(width = 70)

edx[9, ]
```

We can tell this user by 'userId' = 1 rated the movie "Lion King" with 'movieId' = 364 for 5 stars on the date of 'r as.Date(as.POSIXct(edx[9,]$timestamp, origin="1970-01-01"))'. 
The movie "Lion King" was released in 1994, and the movie is categorised as 'r edx[9,]$genres', in total of 5 genres. 


## Methods details

## Final result

## Conclusion

## Refnereces
